---
title: "BME 4550 Assignment 9"
author: "Pallavi Swarup"
date: "October 30, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("limma", version = "3.8")

library(tidyverse)
library(ps)
library(S4Vectors)
library(pkgload)
library(devtools)
install_github("genomicsclass/dagdata")
library(dagdata)
data(admissions)

library(Biobase)
install_github("genomicsclass/GSE5859")
library(GSE5859)
data(GSE5859)



```

# BATCH EFFECTS

```{r}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("S4Vectors", version = "3.8")
```

## Exercises pg 419

### 1. 
```{r}
index = which(admissions$Gender==0)
accepted= sum(admissions$Number[index] * admissions$Percent[index]/100)
applied = sum(admissions$Number[index])
accepted/applied
```

### 2. 
```{r}
atab<-as.data.frame(matrix(0,2,2))
rownames(atab)<-c("Men","Women")
colnames(atab)<-c("Accepted","Rejected")
index = admissions$Gender==1
men=admissions[index,]
women=admissions[!index,]
atab[1,1]= sum(men$Number * men$Percent/100)
atab[1,2]= sum(men$Number*(1-men$Percent/100))
windex = which(admissions$Gender==0)
atab[2,1]= sum(women$Number * women$Percent/100)
atab[2,2]= sum(women$Number * (1-women$Percent/100))
atab
chisq.test(atab)$p.value
```

### 3 & 4. ****************************88
```{r}
major = admissions[1:6,1]
men = admissions[1:6,]
women =admissions[7:12,]
H = (men$Number*men$Percent/100 + women$Number*women$Percent/100) / (men$Number+women$Number)
H
major[which.min(H)]
min(H)
```

### 5.
```{r}
cor(H,men$Number)
cor(men$Number,H)
```

### 6. 
```{r}
cor(H,women$Number)

```

### 7. C

## Exercises pg. 431

### 1. no question (familiarized in console)
```{r}
library(Biobase)
library(GSE5859)
data(GSE5859)
geneExpression= exprs(e)
sampleInfo = pData(e)
```


### 2. 
```{r}
year = format(sampleInfo$date,"%y")
table(year,sampleInfo$ethnicity)
      
tab=table(year,sampleInfo$ethnicity)
print(tab)
x=rowSums(tab!=0)
sum(x>=2)
```
### 3. 

```{r}
month.year = format(sampleInfo$date,"%m%y")
mytab=table(month.year,sampleInfo$ethnicity)
print(mytab)
myx=rowSums(mytab!=0)
mean(myx>=2)
```


#### From this, we can tell that month.year and ethnicity are almost totally confounded, so it is difficult to differentiate whether the outcomes are due to date or ethnicity. this implies that month.year and ethnicity are almost completely confounded. This means that it is hard to separate effects due to date from effects due to our outcome of interest.

### 4. 

```{r}
library(qvalue)
library(genefilter)
year = factor( format(sampleInfo$date,"%y") )
index = which(year%in% c("02","03") & sampleInfo$ethnicity=="CEU")
year = droplevels(year[index])
pval = rowttests(geneExpression[ ,index], year)$p.value
qval = qvalue(pval)
sum(qval$qvalue < 0.05)
```

### 5. 
```{r}
qval$pi0

```

### 6
```{r}
library(qvalue)
library(genefilter)
year = factor( format(sampleInfo$date,"%y") )
index = which(year%in% c("03","04") & sampleInfo$ethnicity=="CEU")
year = droplevels(year[index])
pval = rowttests(geneExpression[ ,index], year)$p.value
qval = qvalue(pval)
sum(qval$qvalue < 0.05)
```
#### Here we can confirm that the processing data has a confounding effect on our measurements

### 7.
```{r}
library(qvalue)
library(genefilter)
ethnicity = factor( sampleInfo$ethnicity)
index = which(ethnicity%in% c("CEU","ASN"))
ethnicity = droplevels(ethnicity[index])
pval = rowttests(geneExpression[ ,index], ethnicity)$p.value
qval = qvalue(pval)
sum(qval$qvalue < 0.05)
```
 

### 8. 

```{r}
library(qvalue)
library(genefilter)
ethnicity = factor( sampleInfo$ethnicity)
year = factor( format(sampleInfo$date,"%y") )
index = which(ethnicity%in% c("CEU","ASN") & year=="05")
ethnicity = droplevels(ethnicity[index])
pval = rowttests(geneExpression[ ,index], ethnicity)$p.value
qval = qvalue(pval)
sum(qval$qvalue < 0.05)
```
```{r}
table(sampleInfo$ethnicity[index])
```


#### There exists a dramatic drop in the number of genes with a q-value<0.05 when the year is fixed. However the sample size also decreased

### 9.

```{r}
library(qvalue)
library(genefilter)
ethnicity = factor( sampleInfo$ethnicity)
year = factor( format(sampleInfo$date,"%y") )
indexASN = which(ethnicity%in% c("ASN") & year=="05")
indexCEU = which(ethnicity%in% c("CEU") & year=="02")
set.seed(3)
indexCEU<-sample(indexCEU,3)
index<-c(indexASN,indexCEU)
ethnicity = droplevels(ethnicity[index])
pval = rowttests(geneExpression[ ,index], ethnicity)$p.value
qval = qvalue(pval)
sum(qval$qvalue < 0.05)
```

## Exercises Pg 442

```{r}
library(GSE5859Subset)
data(GSE5859Subset)
sex = sampleInfo$group
month = factor( format(sampleInfo$date,"%m"))
table( sampleInfo$group, month)
```

### 1. 

```{r}
library(qvalue)
library(genefilter)
sex = factor( sex)
pval = rowttests(geneExpression, sex)$p.value
qval = qvalue(pval)
sum(qval$qvalue < 0.1)
```

### 2. 

```{r}
library(qvalue)
library(genefilter)
sex = factor( sex)
pval = rowttests(geneExpression, sex)$p.value
qval = qvalue(pval)
qvals = qval$qvalues
index = geneAnnotation$CHR[qvals<0.1]%in%c("chrX","chrY")
mean(index)
```

### 3.**********

### 4. 

```{r}
library(qvalue)
library(genefilter)
sex = factor( sex)
pval = rowttests(geneExpression, sex)$p.value
qval = qvalue(pval)
qvals = qval$qvalues
index = which(qvals<0.1 & !geneAnnotation$CHR%in%c("chrX","chrY"))
month = factor( format(sampleInfo$date,"%m"))
pval = rowttests(geneExpression[index,], month)$p.value
mean(pval<0.05)
```

### 5. D: 'X = model.matrix(~sex+month)

### 6. 

```{r}
X = model.matrix(~sex+month)
i = 234
y = geneExpression[i,]
fit = lm(y~X-1)
summary(fit)$coef
```

```{r}
X = model.matrix(~sex+month)
pvals <- t( sapply(1:nrow(geneExpression),function(j){
  y <- geneExpression[j,]
  fit <- lm(y~X-1)
  summary(fit)$coef[2,4] # the 2 picks out the second (sex) coef, the 4 picks out the p value
} ) )
library(qvalue)
library(genefilter)
qval = qvalue(pvals)
qvals = qval$qvalues
sum(qvals<0.1)
```

### 7. 

```{r}
index = geneAnnotation$CHR[qvals<0.1]%in%c("chrX","chrY")
mean(index)
```

### 8. ******

### 9.

```{r}
X = model.matrix(~sex+month)
pvals <- t( sapply(1:nrow(geneExpression),function(j){
  y <- geneExpression[j,]
  fit <- lm(y~X-1)
  summary(fit)$coef[3,4] # third coef, for month
} ) )
library(qvalue)
library(genefilter)
qval = qvalue(pvals)
qvals = qval$qvalues
sum(qvals<0.1)
```


## Exercises on Pg 448

```{r}
library(Biobase)
library(GSE5859Subset)
data(GSE5859Subset)
```

### 1. B: y (1 1 , 1 -1)

### 2. C: D are the coordinates of the prjection U^T Y

### 3. 

```{r}

y = geneExpression - rowMeans(geneExpression)
library(RColorBrewer)
library(rafalib)
mypar(1,2)
n=ncol(y)
cors=cor(y)
cols=colorRampPalette(rev(brewer.pal(11,"RdBu")))(100)
image(1:n,1:n,cors,xaxt="n",yaxt="n",col=cols,xlab="",ylab="",zlim=c(-1,1))
o <- order(sampleInfo$date)
yo<-y[,o]
cors=cor(yo)
cols=colorRampPalette(rev(brewer.pal(11,"RdBu")))(100)
image(1:n,1:n,cors,xaxt="n",yaxt="n",col=cols,xlab="",ylab="",zlim=c(-1,1))
```

#### D: The fact that in the plot ordered by month we see two groups mainly driven by month, and within these we see subgroups driven by date, seems to suggest date more than month per se are the hidden factors.

### 4. 

```{r}
s <- svd(y)
What <- s$v[,1:2] # these are the first two PCs.
rownames(What)<-colnames(y)
round(What,2)
fit = s$u[,1:2]%*% (s$d[1:2]*t(What))
var(as.vector(fit))/var(as.vector(y))
```

### B: pcs = svd(y)$v[,1:2]

### 5. 

```{r}
month=format(sampleInfo$date, "%m")
pcs = svd(y)$v[,1:2]
o = order(sampleInfo$date)
cols = as.numeric(month)[o]
mypar(2,1)
for(i in 1:2){
  plot(pcs[o,i],col=cols,xaxt="n",xlab="")
  label = gsub("2005-","",sampleInfo$date[o])
  axis(1,1:ncol(y),label,las=2)}
```
#### June 23 and June 27

### 7. 

```{r}
s = svd(y)
varexplained = s$d^2/ sum(s$d^2)
plot(varexplained)
sum(varexplained>0.10)
```
#### 2 PCs
 
### 7.
 
```{r}
s = svd(y)
month = factor( format(sampleInfo$date,"%m"))
cors = cor( as.numeric(month),s$v)
plot(t(cors))
which.max(abs(cors))
max(abs(cors))
```

### 8. 0.8297915

### 9. 

```{r}
s = svd(y)
sex = sampleInfo$group
cors = cor( as.numeric(sex),s$v)
plot(t(cors))
which.max(abs(cors))
max(abs(cors))
```

### 10. 0.6236858


### 12. 

```{r}
index = geneAnnotation$CHR[qvals<0.1]%in%c("chrX","chrY")
mean(index)
```

### ALL OF THEM!


## Exercises pg 460


```{r}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("sva", version = "3.8")
library(sva)
```

### 1.

```{r}

s <- svd(geneExpression-rowMeans(geneExpression))
cor(sampleInfo$group,s$v[,1])
sex = sampleInfo$group
mod = model.matrix(~sex)
svafit = sva(geneExpression,mod)
head(svafit$sv)
for(i in 1:ncol(svafit$sv)){
  print( cor(s$v[,i],svafit$sv[,i]) )
}

```


### 2. 

```{r}
index = geneAnnotation$CHR[qvals<0.1]%in%c("chrX","chrY")
mean(index)
```

